Interactive Graphics for Visually Diagnosing Forest Classifiers in R

### 2021-02-01 PPforestVizupdate.Rnw
update
Include comments to work later
update bib with PPforest paper

\section*{Abstract}
This paper describes structuring data and constructing plots to explore forest classification models interactively. A forest classifier is an example of an ensemble since it is produced by bagging multiple trees. The process of bagging and combining results from multiple trees produces numerous diagnostics which, with interactive graphics, can provide a lot of insight into class structure in high dimensions. Various aspects are explored in this paper, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations. The ideas are applied to the random forest algorithm and projection pursuit forest, but could be more broadly applied to other bagged ensembles. Interactive graphics are built in \pkg{R} using the \pkg{ggplot2}, \pkg{plotly}, and \pkg{shiny} packages.

@@

This paper describes structuring interactive graphics to facilitate visual exploration of ensemble classifiers using RFs and projection pursuit forests (PPF)~\citep{dasilvappforest} as examples. The PPF algorithm builds on the projection pursuit tree (PPtree)~\citep{lee2013pptree} algorithm which uses projection pursuit at each tree node to find the best linear combination of variables to separate the classes. The visualization approach is consistent with the framework in \cite{wickham2015visualizing}, and the implementation is built on the newest interactive graphics available in \pkg{R}\citep{RCore}. The purpose is to provide readily available tools for users to explore and improve ensemble fits and obtain an intuition for the underlying class structure in data. Interactive plots are a key component for model visualization that help the user see multivariate relationships and be more efficient in the model diagnosis. Multiple levels of data are constructed for exploration: observation, model and ensemble summaries.

The paper is structured as follows.  Section \ref{key} describes the ensemble components to be accessed.  Section \ref{visen} maps the ensemble components to the visual elements. The web app is described in \ref{app} and further work is discussed in section \ref{fur}.

\section{Diagnostics in forest classifiers}\label{key}

@@

\section{Mapping ensemble diagnostics to visual components}~\label{visen}

@@

\subsection{Individual models: PPtree}

The PPF is composed of individual projection pursuit trees. Figure \ref{pptree} shows a visual ensemble of plots of a tree model on the crab data. There are three nodes for the four class problem. The nodes of this tree are based on projections of the data; the coefficients of which form the building block to calculate the variable importance. The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
The package \pkg{PPtreeViz} provides visual tools to diagnose a PPtree model. The PPF builds on these and modifies a little. The PPtree model is simpler than a regular classification tree because the classes are mostly separated by combinations of variables -- just three projections are needed to see the differences between the four classes.

@@

\subsection{Variable importance}

@@

\subsection{Similarity of cases}

@@

Multidimensional scaling (MDS) is used to reduce the dimension of this matrix, to view the similarity between observations. MDS transforms the data set into a low-dimensional space where the distances are approximately the same as in the full $n$ dimensions. With $G$ groups, the low-dimensional space should be no more than $G-1$ dimensions. Figure \ref{prox1} shows the MDS plots for the 3D space induced by the four groups of the crab data. Color indicates the true species and sex. For this data two dimensions are enough to see the four groups separated quite well. Some crabs are clearly more similar to a different group, especially in examining the sex differences.

@@

\subsection{Uncertainty of cases}

@@

\section{Interactive web app }\label{app}

@@

\subsection{Individual cases}

@@

\subsection{Models}

@@

\subsection{Performance comparison}

@@

The  ROC is used to summarize the trade-off between sensitivity and specificity. The plot shows the sensitivity and specificity when a parameter of classifier is varied \citep{trevor2011elements}. The specificity and sensitivity was computed with the \pkg{pROC} package.
If more than two classes are available a multi-class ROC analysis is needed. Several solutions have been proposed for multi-class ROC. Some of the proposed reduced the multi-class problem to a set of binary problems. The approach used for a multi-class ROC analysis in this paper is called one-against-all ~\citep{allwein2000reducing}.

>>>>

\section*{Abstract}
This article describes structuring data and constructing plots to explore forest classification models interactively. A forest classifier is an example of an ensemble since it is produced by bagging multiple trees. The process of bagging and combining results from multiple trees produces numerous diagnostics which, with interactive graphics, can provide a lot of insight into class structure in high dimensions. Various aspects are explored in this article, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations. The ideas are applied to the random forest algorithm and projection pursuit forest, but could be more broadly applied to other bagged ensembles. Interactive graphics are built in \pkg{R} using the \pkg{ggplot2}, \pkg{plotly}, and \pkg{shiny} packages.

@@

This article describes structuring interactive graphics to facilitate visual exploration of ensemble classifiers using RFs and projection pursuit random forests (PPF)~\citep{da2021projection} as examples. The PPF algorithm builds on the projection pursuit tree (PPtree)~\citep{lee2013pptree} algorithm which uses projection pursuit at each tree node to find the best linear combination of variables to separate the classes. The visualization approach is consistent with the framework in \cite{wickham2015visualizing}, and the implementation is built on the newest interactive graphics available in \pkg{R}\citep{RCore}. The purpose is to provide readily available tools for users to explore and improve ensemble fits and obtain an intuition for the underlying class structure in data. Interactive plots are a key component for model visualization that help the user see multivariate relationships and be more efficient in the model diagnosis. Multiple levels of data are constructed for exploration: observation, model and ensemble summaries.

% Should I include some references about interpretable machine learning

The article is structured as follows.  Section \ref{key} describes the ensemble components to be accessed.  Section \ref{visen} maps the ensemble components to the visual elements. The web app is described in \ref{app} and further work is discussed in section \ref{fur}.

\section{Diagnostics in Forest Classifiers}\label{key}

@@

\section{Mapping Ensemble Diagnostics to Visual Components}~\label{visen}

@@

\subsection{Individual Models: PPtree}

The PPF is composed of individual projection pursuit trees. Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data. There are three nodes for the four class problem. The nodes of this tree are based on projections of the data; the coefficients of which form the building block to calculate the variable importance. The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
The package \pkg{PPtreeViz} provides visual tools to diagnose a PPtree model. The PPF builds on these and modifies a little. The PPtree model is simpler than a regular classification tree because the classes are mostly separated by combinations of variables -- just three projections are needed to see the differences between the four classes.

@@

\subsection{Variable Importance}

@@

\subsection{Similarity of Cases}

@@

Multidimensional scaling (MDS) is used to reduce the dimension of this matrix, to view the similarity between observations. MDS transforms the dataset into a low-dimensional space where the distances are approximately the same as in the full $n$ dimensions. With $G$ groups, the low-dimensional space should be no more than $G-1$ dimensions. Figure \ref{prox1} shows the MDS plots for the 3D space induced by the four groups of the crab data. Color indicates the true species and sex. For this data two dimensions are enough to see the four groups separated quite well. Some crabs are clearly more similar to a different group, especially in examining the sex differences.

@@

\subsection{Uncertainty of Cases}

@@

\section{Interactive Web App }\label{app}
%Todo the PPforestViz pakage

@@

\subsection{Individual Cases}

@@

\subsection{Models}
%I have to modify the selecction of the tree instead to select the tree based on the importance variable will be better select the tree based on the ooberror. Move the boxplot and select using thisone.
%Include some alternative visualization fro Random forest
%This is the oly tabset that can not be reproduced using RandomForest because we don't have the same information than in PPforest. Also doesn't make sense present the 1D projection for each of the node partitions.

@@

\subsection{Performance Comparison}

@@

The  ROC is used to summarize the trade-off between sensitivity and specificity. The plot shows the sensitivity and specificity when a parameter of classifier is varied \citep{trevor2011elements}. The specificity and sensitivity was computed with the \pkg{pROC} package.
If more than two classes are available a multi-class ROC analysis is needed. Several solutions have been proposed for multi-class ROC. Some of the proposed reduced the multi-class problem to a set of binary problems. The approach used for a multi-class ROC analysis in this article is called one-against-all ~\citep{allwein2000reducing}.
